\blue{...Numerical investigation of viscous flow instabilities through heterogeneous porous media using adaptive and unstructured meshes under different mobility scenarios...}

### 2017-08-10 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
latest updates

\subsection{3-D channel flows} \label{section:results_3D}

The last of the simulations introduces a $3$D domain that will showcase the idea of the preferential flow path and finger growth using different permeability regions, as these are decribed in figure \ref{fig:3DChannel_PermMesh}. In large scale geological formations preferential flow allows much faster contaminant transport and creates significant consequences for ground-water quality.

\medskip
As already mentioned in section \ref{section:ViscousInstabilities}, \textit{viscous fingering} is characterized by the formation of flow patterns, called viscous fingers. The geometrical evolution of these patterns provides good indications on capturing and tracking any discontinuities or instabilities in the flow front. Figure \ref{fig:3DChannel_PermMesh} describes the permeability pattern as this is implied in a $3D$ domain. The surface on the left hand side of the domain is the area where fluid $1$ is injected with a velovity $ u=2.5 cm/sec $. No-slip conditions are implied around the domain (\ie insulation, no exchange of material or Neumann boundary condition for mass transfer) while at the right hand side of the domain is the outlet. The control volume is constructed in such a way so as to be able to focus and capture the preferential path of flow as this can be seen from figures \ref{fig:3DChannel_planes}-\ref{fig:3DChannel_Isosurf}.

\medskip
Figure \ref{fig:3DChannel_planes} gives a comprehensive description of the contours of phase $1$ darcy velocity, during the beggining and just before the end of the simulation. While the saturation level for phase $1$, remains between the values of $0.25-0.75$ the main difference between the figures \ref{fig:3DChannel_planes} (a) and (b) is the intensity of the flow as this can be given by the darcy velocity contours. In the beginning of the simulation most of the contours for phase $1$ are populated and achieve their higher values closer to the inlet region. At this point a comparison between the figures \ref{fig:3DChannel_planes}(a)-\ref{fig:3DChannel_sat}(a) and \ref{fig:3DChannel_planes}(b)-\ref{fig:3DChannel_sat}(f) as they both describe the darcy velocity contours and the saturation front for the same timesteps, is necesserary.

\medskip
In previous sections \ref{section:results:setup}-\ref{section:results_hete_fix_adapt}, a $2$D domain was used to describe one or more finger formations. These flow instabilities were triggerded because of the permeability differences in the domain. Using the $3D$ domain, the difference between a flow instability and the preferential flow path that may include smaller fingers is more obvious. Figures \ref{fig:3DChannel_sat}(a)-(f) give a description of the preferable flow path in various time-steps alongside with the progression of the saturation front that now becomes more obvious. Based on the boundary conditions that implemented, the preferencial path of the flow can been seen. Heterogeneity in the reservoir controls the spatial distribution of porosity and permeability. The permeability depends on the number, size and connectivity of the pores. An on purpose implementation of significant high permeability zones (compared with the lower ones) also know as higher permeability channels, allows the domain to give a good approximation of the so called "thief zones".

\medskip 
The saturation front progress and the idea of preferential flow path is followed by a more detail representation as this can be found in figure \ref{fig:3DChannel_Isosurf}. For the same time-steps an inner view on the domain provides with a more detailed description of not only the flow path itself as well as the minor fingers that may be created. At this point resolution plays an important role on capturing these patterns in detail and is also a limititation that needs to be overcome.

\medskip
A quick review to the mechanisms of viscous fingering (\ie shielding, spreading and tip-splitting, \citet{homsy_1987}) is required. In all the $2$D previous cases (sections \ref{section:results:setup}-\ref{section:results_hete_fix_adapt}), as interfacial surface tension (IFT) becomes weak, the front of the steady finger is susceptible to a viscous-fingering instability. Through \textit{shielding} mechanism, one of these fingers will eventually outgrow the other and, will then spread (\textit{spreading}) to occupy the appropriate width of the domain ahead of it. Following this process, it reaches a width that is again unstable and the pattern repeats. So far shielding and spreading were dominating the flow but, using the current $3$D domain, the conditions and the mechanics of tip-splitting along a front can be qualitatively understood.

\section{Conclusions}\label{Section:Conclusion}

@@

\medskip
As usually seen for high viscosity contrast, there is a leading finger (tip splitting mechanism) that breaks out of the front by suppressing its growth to the point that smaller fingers will break out driven by surface tension. At long times the large finger advances approximately at in a constant flow rate. In order to run all these simulations in detail, the use and benefits of dynamic mesh adaptivity to capture the viscous fingering in reservoir flows was introduced. Commercial simulators provide the user with option of an unstructured type of mesh still they are not able to dynamically adapt the mesh as the simulation develops and they increase the computational cost as the saturation front changes slowly.

\medskip
In summary, the method that has been described in this paper as well as the simulations performed have a number of distinct advantages: 
\begin{enumerate}
\item[1.] The geological model can be designed from the beginning without any previous reference or geometrical limitations due to both the flexibility of method and to the fact that the geological characteristic and parameters can be easily introduced.
\item[2.] Computational cost can be reduced because the complex geometries and surfaces are less expensive to generate and manage using fewer grid cells. The basis of the this argument is the flexibility and variance of the unstructured grid in time and space.
\item[3.] Production simulations run directly on the geological model without the need to upscale.\\
\end{enumerate}

>>>>

%%%%%%%%%%%%%
\subsection{3-D channel flows} \label{section:results_3D}

The last of the simulations introduced is a $3$D domain that will showcase the idea of the preferential flow path and finger growth using different permeability regions, as these are decribed in figure \ref{fig:3DChannel_PermMesh}. In large scale geological formations preferential flow allows much faster contaminant transport and creates significant consequences for ground-water quality.

\medskip
As already mentioned in section \ref{section:ViscousInstabilities}, \textit{viscous fingering} is characterized by the formation of flow patterns, called viscous fingers. The geometrical evolution of these patterns provides good indications on capturing and tracking any instabilities in the flow front. Figure \ref{fig:3DChannel_PermMesh} describes the permeability pattern as this is implied in a $3D$ domain. The surface on the left hand side of the domain is the area where fluid $1$ is injected with a velovity $ u=2.5 cm/sec $. No-slip conditions are implied everywhere else (\ie insulation, no exchange of material or Neumann boundary condition for mass transfer) while at the right hand side of the domain there is the outlet. The control volume is constructed in such a way so as to be able to focus and capture the preferential path of flow as this can be seen from figures \ref{fig:3DChannel_planes}-\ref{fig:3DChannel_Isosurf}.

\medskip
Figure \ref{fig:3DChannel_planes} gives a comprehensive description of the contours of phase $1$ darcy velocity, in times related to the beginning and just before the end of simulation. While the saturation level for phase $1$, remains between the values of $0.25-0.75$ the main difference between the figures \ref{fig:3DChannel_planes} (a) and (b) is the intensity of the flow as this can be given by the darcy velocity contours. In the beginning of the simulation most of the contours for phase $1$ are populated and achieve their higher values closer to the inlet region. At this point a comparison between the figures \ref{fig:3DChannel_planes}(a)-\ref{fig:3DChannel_sat}(a) and \ref{fig:3DChannel_planes}(b)-\ref{fig:3DChannel_sat}(f) as they both describe the darcy velocity contours and the saturation front for the same timesteps, is necesserary.

\medskip
In sections \ref{section:results:setup}-\ref{section:results_hete_fix_adapt}, a $2$D domain was used to describe one or more finger formations. These flow instabilities were triggerded from the permeability differences in the domain. Using the $3D$ domain, the difference between a flow instability and the preferential flow path that may include smaller fingers is more obvious. Figures \ref{fig:3DChannel_sat}(a)-(f) describe the flow path in various time-steps alongside with the progression of the saturation front that now becomes more obvious. Heterogeneity in the reservoir controls the spatial distribution of porosity ($\phi =0.2$ in the current case) and permeability. The permeability depends on the number, size and connectivity of the pores. An on purpose implementation of significant high permeability zones (compared with the lower ones) also know as higher permeability channels, allows the domain to give a good approximation of the so called "thief zones".

\medskip 
The saturation front progress and the idea of preferential flow path is followed by a more detail representation as this can be found in figure \ref{fig:3DChannel_Isosurf}. In figures \ref{fig:3DChannel_sat_adapt}(a)-(f) an unstructured and adaptive mesh has been introduced in order to achieve higher resolution and focus the computational effort where is needed in he domain. For the same time-steps an inner view on the domain provides with a more detailed description of not only the flow path itself as well as the minor fingers that may be created. At this point resolution plays an important role on capturing these patterns in detail and is also a limititation that needs to be overcome.

\medskip
A quick review to the mechanisms of viscous fingering (\ie shielding, spreading and tip-splitting, \citet{homsy_1987}) is required. In all the $2$D previous cases (sections \ref{section:results:setup}-\ref{section:results_hete_fix_adapt}), as interfacial surface tension (IFT) becomes weak, the front of the steady finger is susceptible to a viscous-fingering instability. Through \textit{shielding} mechanism, one of these fingers will eventually outgrow the other and, will then spread (\textit{spreading}) to occupy the appropriate width of the domain ahead of it. Following this process, it reaches a width that is again unstable and the pattern repeats. So far shielding and spreading were dominating the flow but, using the current $3$D domain and by applying two types of meshes, despite the resolution restrictions, the conditions and the mechanics of tip-splitting along a front can be qualitatively understood.

%%%%%%%%%%%%%
\section{Conclusions}\label{Section:Conclusion}

@@

\medskip
As seen for high viscosity contrast, there is a leading finger (\textit{tip splitting mechanism}) that breaks out of the front by suppressing its growth to the point that smaller fingers will break out driven by surface tension. At long times the large finger advances approximately at in a constant flow rate. In order to run all these simulations in detail, the use and benefits of dynamic mesh adaptivity to capture the viscous fingering in reservoir flows was introduced for all the $2D$ and $3D$ cases. Commercial simulators provide the user with option of an unstructured type of mesh, still they are not able to dynamically adapt the mesh where is needed and manage the computational cost more efficient.

\medskip
In summary, the method that has been described in this paper as well as the simulations performed have a number of distinct advantages: 
\begin{enumerate}
\item[1.] The geological model can be designed from the beginning without any previous reference or geometrical limitations due to both the flexibility of method and to the fact that the geological characteristic and parameters can be easily introduced.
\item[2.] Computational cost can be reduced, because the complex geometries and surfaces are less expensive to generate and manage using fewer grid cells. The basis of the this argument is the flexibility and variance of the adaptive and unstructured grid in time and space.
\item[3.] Production simulations run directly on the geological model without the need to upscale.\\
\end{enumerate}
